---
title: "Forecasting Heart Disease Risks"
output: 
  html_document:
    theme:
      bg: "#1b1b1b"
      fg: "#FDF7F7" 
      primary: "#98002E"
      base_font:
        google: Montserrat
      code_font:
        google: JetBrains Mono
    highlight: espresso
    css: "resources/styles.css"
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Set default CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

#Install the following libraries if required
# List of required packages
required_pkgs <- c("tidyverse", "RCurl", "caret", "randomForest",
                   "e1071", "rpart", "corrplot", "GGally", "pROC")

# Install and load each package
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

```

## Introduction



<br>

## Business Understanding
### Problem statement
To develop models for an insurance company using the Heart Disease dataset from the UCI Machine Learning Repository. The goal is to predict the likelihood of a person developing heart disease, which would help the insurance company estimate health risks and adjust premiums accordingly.


<br>

## Data Understanding
The dataset contains various features related to patients' health and demographic information. We will explore the dataset to understand its structure and relationships between variables.

### Data dictionary
The dataset contains 14 key attributes that are either numerical or categorical.

| Attribute	| Type	| Description	| Constraints/ Rules |
|:----------|:------|:------------|:-------------------|
| `age`	| Numerical	| The age of the patient in years	| Range: 29-77 (based on dataset statistics) |
| `sex`	| Categorical	| The gender of the patient	| Values: 1 = Male, 0 = Female |
| `cp` | Categorical	| Type of chest pain experienced by the patient | 	Values: 1 = Typical angina, 2 = Atypical angina, 3 = Non-anginal pain, 4 = Asymptomatic |
| `trestbps`	| Numerical	| Resting blood pressure of the patient, measured in mmHg	| Range: Typically, between 94 and 200 mmHg |
| `chol` | Numerical	| Serum cholesterol level in mg/dl | Range: Typically, between 126 and 564 mg/dl |
| `fbs`	| Categorical	| Fasting blood sugar level > 120 mg/dl | 	Values: 1 = True, 0 = False |
| `restecg` | Categorical	| Results of the patient’s resting electrocardiogram	| Values: 0 = Normal, 1 = ST-T wave abnormality, 2 = Probable or definite left ventricular hypertrophy |
| `thalach`	| Numerical	| Maximum heart rate achieved during a stress test | Range: Typically, between 71 and 202 bpm |
| `exang`	| Categorical	| Whether the patient experiences exercise-induced angina	| Values: 1 = Yes, 0 = No |
| `oldpeak`	| Numerical	| ST depression induced by exercise relative to rest (an ECG measure)	| Range: 0.0 to 6.2 (higher values indicate more severe abnormalities) |
| `slope`	| Categorical	| Slope of the peak exercise ST segment	| Values: 1 = Upsloping, 2 = Flat, 3 = Downsloping |
| `ca` | Numerical	| Number of major vessels colored by fluoroscopy	| Range: 0-3 |
| `thal` | Categorical	| Blood disorder variable related to thalassemia	| Values: 3 = Normal, 6 = Fixed defect, 7 = Reversible defect |
| `target` | Categorical	| Diagnosis of heart disease	| Values: 0 = No heart disease, 1 = Presence of heart disease |


<br>

## Data Preparation
### Data loading
Load the dataset from the UCI website to memory
```{r}
# Load the dataset
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"

# Read the dataset into a dataframe
Heart.df <- read.csv(text = getURL(url), header = FALSE, na.strings = "?")
```

Rename the columns into a meaningful column names
```{r}
colnames(Heart.df) <- c("age", "sex", "cp", "trestbps", "chol", "fbs",
                        "restecg", "thalach", "exang", "oldpeak", 
                        "slope", "ca", "thal", "target")
```

Display dimensions of the dataset
```{r}
dim(Heart.df)
```

Display the first six rows of the dataset
```{r}
head(Heart.df)
```

Display the structure of the dataframe
```{r}
str(Heart.df)
```

Display the statistical summary of the dataframe
```{r}
summary(Heart.df)
```

### Data preprocessing
We will preprocess the data by handling missing values, encoding categorical variables, and scaling numerical features.


<br>

## Modeling



<br>

## Evaluation



<br>

## Deployment


<br>

## Conclusion

